---
title: "What gives better mpg, Automatic or Manual transmission?"
author: "Nitin Aggarwal"
date: "September 27, 2015"
output: pdf_document
fontsize: 10pt
---

## Executive Summary

To answer the dependence of mpg on Automatic vs Manual transmission, `mtcars` data is studied. After exploration a few models are derived and a best model is chosen from which it was found that the dependence of `mpg`, miles/gallon for different cars is confounded on `hp`, `wt`, and `cyl`, i.e. the horsepower, weight, and number of cylinders. The final linear models shows that the Manual cars have **1.809MPGS** more than that of Automatic cars. 


## Introduction

We are working with the `mtcars` data which can be loaded through `datasets` as follows:

```{r}
data("mtcars")
```

```{r, echo = FALSE}
options(digits = 3)
```
Let us also convert the qualitative features as factor variables.

```{r, echo = FALSE}
mtcars$cyl<-as.factor(mtcars$cyl)
mtcars$vs<-as.factor(mtcars$vs)
mtcars$am<-as.factor(mtcars$am)
mtcars$gear<-as.factor(mtcars$gear)
```

A call to the R help shows the following description to the data:

*The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).*

#### Goal
As the title suggests the main goal of this analysis is to study the effect of type of transmission, automatic vs manual, on mpg. 


## Exploration

In **Appendix 1** we have a sequence of plots ploting `mpg` against other features with respect to transmission type. 

Let us also get the correlations of different features present in `mtcars` as follows:

```{r}
ind<-sapply(1:11,function(i)!is.factor(mtcars[1:nrow(mtcars),i]))
mtcars.corr<-cor(mtcars[,ind])
mtcars.corr
```

From these one can infer that `mpg` has high level of positive or negative correlation with features in columns `disp`, `hp`, `wt`, `dart`, `carb`. Of these `disp`, `quesc`, `carb` shows high correlation with `hp`, giving a possibility of them being redundant. Also, from the graphs in **Appendix 1** we notice that among the factor columns, `cyl`, `vs`, and `gear`, all of them are shown to affect the `mpg`, but the columns `cyl` and `gear` are also showing interaction with `am` column. 

## Linear Regression

We will fit the following models we found from our analysis, and choose the best of them.

1. `mpg` against just `am`

2. `mpg` against `am`, `cyl`

3. `mpg` against `am`, `cyl`, `gear`

4. `mpg` against interactions between `am` and `cyl`

5. `mpg` against interactions between `am` and `gear`

6. `mpg` against interactions between `am` and `cyl`, and between `am` and `gear`

7. `mpg` against `hp` with best of above models.

8. above model with `wt` included

Let us create the first six models first:

```{r, echo = FALSE}
fit<-list()

fit[[1]] <- lm(mpg ~ am, mtcars)

fit[[2]] <- lm(mpg ~ am + cyl, mtcars)

fit[[3]] <- lm(mpg ~ am + cyl + gear, mtcars)

fit[[4]] <- lm(mpg ~ am * cyl, mtcars)

fit[[5]] <- lm(mpg ~ am * gear, mtcars)

fit[[6]] <- lm(mpg ~ am * cyl + am * gear, mtcars)

```

Let us choose the best of top three first.

```{r}
anova(fit[[1]], fit[[2]], fit[[3]])
```

Choosing the second model, let us compare it with the fourth and sixth model, ignoring fifth because of absense of `cyl` in it.

```{r}
anova(fit[[2]], fit[[4]], fit[[6]])
```

From this result because of high p values we keep the second of above six model and compare it with the updated by `hp` and then that with `wt`. These models are tested against the second model above in one line as follows:

```{r}
anova(fit[[2]],
      update(fit[[2]], .~. + hp),
      update(fit[[2]], .~. + hp + wt))

```

From this we choose the third of these models. Following presents the details about this model:

```{r}
fitf<-update(fit[[2]], .~. + hp + wt)
fitf
```

Let us now quantify this plot using a diagnosis using residuals:

## Residuals and diagnositcs

The residuals and the Normal Q-Q plot in `Appendix 2` shows that the data is free of heteroscedasticity and is normal. 


## Conclusion

We get following two results from our analysis:

1. Manual transmission gives more mileage per gallon given the fixed values of cyl, hp, and wt. 

2. Manual transmission increases the mileage by `r fitf$coefficients[2]` miles per gallon. 

## Appendix 1


```{r, echo = FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r echo = FALSE}
library(ggplot2)
g<-list()
for(i in c(2:8,10:11)){
  
  if(is.factor(mtcars[,i])){
  
    g[[length(g)+1]]<-ggplot(data = data.frame(y = mtcars[,1], x = mtcars[,i], type = mtcars[,9]), 
           aes(y = y, colour = type, x = x))+ 
      geom_point() + 
      xlab(colnames(mtcars)[i]) + theme(legend.position="none") +
      ylab("mpg") + 
      ggtitle(paste("mpg vs ", colnames(mtcars)[i], " given transmission."))
  } else{
      g[[length(g)+1]]<-ggplot(data = data.frame(y = mtcars[,1], x = mtcars[,i], type = mtcars[,9]), 
           aes(y = y, colour = type, x = x))+ 
        geom_point() + 
        xlab(colnames(mtcars)[i]) + theme(legend.position="none") +
        ylab("mpg") + 
        ggtitle(paste("mpg vs ", colnames(mtcars)[i], " given transmission."))
        geom_violin()  
  }
}
multiplot(g[[1]],g[[2]], g[[3]],g[[4]],cols = 2)
multiplot(g[[5]],g[[6]], g[[7]],g[[8]],cols = 2)
g[[9]]

```


## Appendix 2

```{r, echo = FALSE}
plot(fitf, which = c(1,2))
```
